WORCS: A Workflow for Open Reproducible Code in Science

### 2020-11-03 paper/Manuscript_DS.Rmd
R1 C6 Dependency management

For analyses conducted entirely in `R`, the recently released package `renv` [@usheyRenvProjectEnvironments2020]
strikes a great balance between user-friendliness and robustness.
Developed by the team behind RStudio, 
this package simply maintains a text-based,
human-readable log of all packages used, their version numbers, and where they
were installed from (e.g., CRAN, Bioconductor, GitHub).
This lightweight approach safeguards reproducibility as long as these repositories maintain their archives.
The text-based log file can be version controlled with Git.
The `renv` function `snapshot()` is used to add packages to the log file.
This function identifies all packages loaded using `library()` calls in the scripts in the project directory.

All solutions to computational reproducibility require users to make their dependencies explicit, often by reinstalling all software for each project in a controlled environment.
This can lead to long installation times and large memory requirements.
Similarly, `renv` requires users to call `install.packages()` for each library used - but it does not reinstall packages for each new project.
Instead, packages reside in a cache that is shared between all projects on the computer.
If a project uses a package that is already in the cache, it is not reinstalled, but linked from the cache.
This improves the user experience by saving time and memory.
When someone else loads your project, `renv` will install all of the required packages.
As opposed to alternative solutions, `renv` is tightly integrated with the
native R package management system.
Therefore, it does not require the users to change their workflow of installing/removing and using packages, and requires no technical skill to setup.
It is not a fail-safe method to ensure strict computational reproducibility,
but it meets most reproducibility requirements using only tools
native to RStudio, and is therefore easier to adopt for `R` users.

If stricter reproducibility is required,
users might choose to conduct all analyses in a "Docker container": An environment that behaves like a
virtual computer, which can be stored like a time capsule, and
identically reinstated on a user computer, or in the cloud.
This is effective, and also preserves software outside the `R` environment (e.g.,
preprocessing tools for neuroimaging data).
The [stetup with Docker vignette](https://cjvanlissa.github.io/worcs/articles/docker.html)
describes how to instantiate a Docker build running `worcs` and its dependencies.
For more information on this topic,
@peikertReproducibleDataAnalysis2019 describe 
a Docker-based workflow for computational reproducibility that is
conceptually and practically compatible with WORCS.
One remaining challenge of Docker is that it is somewhat difficult to set up for novice users.
To overcome this challenge, an `R` package is currently in development to facilitate Docker integration [`repro`, @repro].
Another 
user-friendly Docker-based workflow is offered by the cloud-based collaboration
platform 'Code Ocean'. Although Code Ocean enjoys some popularity among 
scientists, it is not clear whether this platform
is GDPR compliant - as of this writing, its [terms of use](https://codeocean.com/terms-of-use) do not address the GDPR, nor is it clear where
data are stored.
The necessity to upload human participant data is a liability
when using any
cloud-computing solution for data analysis.

>>>>

For dependency management, the `R` implementation of `worcs` relies on the recently released package `renv` [@usheyRenvProjectEnvironments2020].
This package strikes a great balance between user-friendliness and robustness.
Developed by the team behind RStudio, 
`renv` maintains a text-based,
human-readable log of all packages used, their version numbers, and where they
were installed from (e.g., CRAN, Bioconductor, GitHub).
This text-based log file can be version controlled with Git.
When someone else loads your project,
`renv` will install all of the required packages from this list.
This lightweight approach safeguards reproducibility as long as these repositories maintain their archives.
The `worcs` project template in RStudio automatically sets up `renv` for dependency management if the checkbox labeled "renv" is selected during project creation.

All solutions to computational reproducibility require users to make their dependencies explicit, often by reinstalling all software for each project in a controlled environment.
This can lead to long installation times and large memory requirements.
Similarly, `renv` requires users to call `install.packages()` for each library used - but it does not reinstall packages for each new project.
Instead, packages reside in a cache that is shared between all projects on the computer.
If a project uses a package that is already in the cache, it is not reinstalled, but linked from the cache.
This improves the user experience by saving time and memory.
As opposed to alternative solutions, `renv` is tightly integrated with the
native R package management system.
It monitors the scripts in the project directory for `library()` calls,
and notifies the user when it is necessary to update the dependencies file using the `snapshot()` function.
Therefore, `renv` does not require the users to change their workflow of installing/removing and using packages, and requires little technical skill to setup.
It is not a fail-safe method to ensure strict computational reproducibility,
but it meets most reproducibility requirements using only tools
native to `R` and RStudio.

When a project is conducted in a different software environment than `R`,
or if an R-project has external dependencies not tracked by `renv`,
then users might choose to use [Docker](https://www.docker.com/) for dependency management.  
Docker instantiates a "Docker container": An environment that behaves like a
virtual computer, which can be stored like a time capsule, and
identically reinstated on a user computer, or in the cloud.
Although the use of Docker falls outside of the scope of the present paper,
for existing users, a Docker build running `worcs` and its dependencies is available.
The [stetup with Docker vignette](https://cjvanlissa.github.io/worcs/articles/docker.html)
explains how to instantiate this build.
For more information on this topic,
@peikertReproducibleDataAnalysis2019 describe 
a Docker-based workflow for computational reproducibility that is
conceptually and practically compatible with WORCS.

As Docker is somewhat challenging to set up for novice users,
an `R` package is currently in development to facilitate Docker integration [`repro`, @repro].
Another example of a Docker-based workflow is the cloud-based collaboration
platform 'Code Ocean'.
Although Code Ocean enjoys some popularity among 
scientists, it is not clear whether this platform
is GDPR compliant - as of this writing, its [terms of use](https://codeocean.com/terms-of-use) do not address the GDPR, nor is it clear where
data are stored.
The necessity to upload human participant data for analysis is an important consideration and potential liability when using any cloud-computing solution.

### 2020-11-03 paper/reviews_ds.Rmd
R1 C6 Dependency management

class       : "draft"

output      : papaja::revision_letter_pdf
---

@@

\Assignment{CJ}
\WorkInProgress
\Medium

\RC{It is not clear how worcs addresses depedency management. The section lists a series of tools that can be used, but the section is not very specific to worcs.}

Response: Not true; explain that worcs project automatically initializes `renv`.

\RC{The workflow of releasing data products on GitHub is not ideal. First, because datasets of 100 MBs usually require special handling. Second, it doesn't follow very well the FAIR principles for findability and attribution of data. Instead, repositories like Zenodo or FigShare could be user to address data storage, with proper metadata and its corresponding citation. The authors seem to not describe very well what would happen in data science experiments where the data size is significant.}

@@

\Assignment{CJ}
\WorkInProgress
\Each

@@

# \Reviewer \#3

@@

# \Reviewer \#4

>>>>

class       : "draft"
bibliography      : ["worcs.bib"]  
output      : papaja::revision_letter_pdf
---

@@

\RC{C6: It is not clear how worcs addresses depedency management. The section lists a series of tools that can be used, but the section is not very specific to worcs.}

Response: We apologize for the apparent lack of clarity of this section. 
We have restructured and clarified this section to convey the following information:

1) Introduce the general concept of dependency management
2) Introduce `renv`, the solution for dependency management used by the R-implementation of `worcs`
3) Acknowledge the more comprehensive approach to dependency management offered by Docker

We have also, in response to another Reviewer comment, created a Docker Hub build that runs `worcs`. Therefore, as of this revision, both of the tools discussed in this section (i.e., `renv` and Docker) are directly relevant for WORCS-users, and the section should read less like a "series of tools that can be used".
If the Reviewer so desires, we would also be willing to cut the brief discussion of Code Ocean - but this had been included at the request of an informal reviewer of this paper (Daniël Lakens).

\RC{C7: The workflow of releasing data products on GitHub is not ideal. First, because datasets of 100 MBs usually require special handling. Second, it doesn't follow very well the FAIR principles for findability and attribution of data. Instead, repositories like Zenodo or FigShare could be user to address data storage, with proper metadata and its corresponding citation. The authors seem to not describe very well what would happen in data science experiments where the data size is significant.}

BV: Can you please argue for why GitHub is equivalent to Zenodo for this purpose?
AL: Can you address the reference to FAIR principles in relation to GitHub?
CJ: I will address "big" data

@@

@@

# Reviewer \#3

@@

# Reviewer \#4
